---
title: "Bivariate and Multivariate Graphs with Plotly Express"
author: "SI Analytics"
format:
  html:
    css: global/style/si-analytics.css
    toc: true
    toc-location: left
    number-sections: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  error: false
---

## Introduction

In this lesson, we extend beyond univariate plots to explore **bivariate and multivariate visualizations** using Plotly in R within the RStudio environment.

These visualizations help us understand:

- Relationships between two numeric variables
- Distributions of numeric variables across categories
- Relationships between categorical variables
- Trends over time
- Patterns involving three or more variables

---

## Learning Objectives

By the end of this lesson, you should be able to:

- Create scatter plots (numeric vs numeric)
- Create grouped histograms and box/violin plots (numeric vs categorical)
- Create stacked and percent-stacked bar charts (categorical vs categorical)
- Create time series plots
- Create bubble charts
- Use faceting to compare subsets of data

---

## Libraries

```{r}
library(plotly)
library(dplyr)
library(gapminder)
library(reactable)
```

---

# Numeric vs Numeric

We will use the built-in `mtcars` dataset.

```{r}
tips <- read.csv(
  "https://raw.githubusercontent.com/plotly/datasets/master/tips.csv"
)

# Render interactive table
tips %>%
  head() %>%
  reactable(
    bordered = TRUE,
    compact = TRUE
    )
```

## Scatter Plot

```{r}
plot_ly(
  tips,
  x = ~total_bill,
  y = ~tip,
  type = "scatter"
) 
```

Let’s enhance the scatter plot by adding labels and a title.


```{r}
plot_ly(
  tips,
  x = ~total_bill,
  y = ~tip,
  type = "scatter",
  mode = "markers"
) %>%
  layout(
    title = "Relationship Between Total Bill and Tip",
    xaxis = list(title = "Total Bill ($)"),
    yaxis = list(title = "Tip ($)")
  )
```



---

::: {.callout-tip title="Practice"}

### Practice: Horsepower vs MPG

Create a scatter plot showing the relationship between `hp` and `mpg`.

```{r}
# Your code here
```
:::

```{r}
#| include: false
plot_ly(
  mtcars,
  x = ~hp,
  y = ~mpg,
  type = "scatter",
  mode = "markers"
)
```

---

# Numeric vs Categorical

Convert some variables to factors.

```{r}
mtcars <- mtcars %>%
  mutate(
    cyl = factor(cyl),
    gear = factor(gear)
  )
```

## Grouped Histogram

```{r}
plot_ly(
  mtcars,
  x = ~mpg,
  color = ~cyl,
  type = "histogram"
)
```

---

## Violin Plot

```{r}
plot_ly(
  mtcars,
  x = ~cyl,
  y = ~mpg,
  type = "violin",
  box = list(visible = TRUE),
  meanline = list(visible = TRUE)
)
```

---

::: {.callout-tip title="Practice"}

### Practice: MPG by Gear

Create a violin or box plot comparing `mpg` across `gear`.

```{r}
# Your code here
```
:::

```{r}
#| include: false
plot_ly(
  mtcars,
  x = ~gear,
  y = ~mpg,
  type = "box"
)
```

---

# Summary Bar Charts (Mean + SD)

```{r}
summary_df <- mtcars %>%
  group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg)
  )

plot_ly(
  summary_df,
  x = ~cyl,
  y = ~mean_mpg,
  type = "bar",
  error_y = list(array = summary_df$sd_mpg)
)
```

---

# Categorical vs Categorical

## Stacked Bar Chart

```{r}
plot_ly(
  mtcars,
  x = ~cyl,
  color = ~gear,
  type = "histogram"
)
```

---

## Percent Stacked Bar Chart

```{r}
plot_ly(
  mtcars,
  x = ~cyl,
  color = ~gear,
  type = "histogram"
) %>%
  layout(barnorm = "percent")
```

---

::: {.callout-tip title="Practice"}

### Practice: Percent Distribution of Gears by Cylinders

Create a percent-stacked bar chart of `gear` by `cyl`.

```{r}
# Your code here
```
:::

```{r}
#| include: false
plot_ly(
  mtcars,
  x = ~cyl,
  color = ~gear,
  type = "histogram"
) %>%
  layout(barnorm = "percent")
```

---

# Time Series

Using the `gapminder` dataset.

```{r}
head(gapminder)
```

## Nigeria Population Over Time

```{r}
nigeria <- gapminder %>%
  filter(country == "Nigeria")

plot_ly(
  nigeria,
  x = ~year,
  y = ~pop,
  type = "scatter",
  mode = "lines+markers"
)
```

---

::: {.callout-tip title="Practice"}

### Practice: GDP per Capita of Ghana

Create a time series plot for Ghana’s `gdpPercap`.

```{r}
# Your code here
```
:::

```{r}
#| include: false
ghana <- gapminder %>%
  filter(country == "Ghana")

plot_ly(
  ghana,
  x = ~year,
  y = ~gdpPercap,
  type = "scatter",
  mode = "lines+markers"
)
```

---

# Bubble Charts (Three Variables)

```{r}
g_2007 <- gapminder %>%
  filter(year == 2007)

plot_ly(
  g_2007,
  x = ~gdpPercap,
  y = ~lifeExp,
  size = ~pop,
  color = ~continent,
  type = "scatter",
  mode = "markers"
)
```

---

::: {.callout-tip title="Practice"}

### Practice: Bubble Plot of Asia 2007

Filter to Asia and create a bubble plot of GDP vs Life Expectancy sized by population.

```{r}
# Your code here
```
:::

```{r}
#| include: false
asia_2007 <- g_2007 %>% filter(continent == "Asia")

plot_ly(
  asia_2007,
  x = ~gdpPercap,
  y = ~lifeExp,
  size = ~pop,
  type = "scatter",
  mode = "markers"
)
```

---

# Faceting

```{r}
plot_ly(
  g_2007,
  x = ~gdpPercap,
  y = ~lifeExp,
  color = ~continent,
  type = "scatter",
  mode = "markers"
) %>%
  layout(title = "GDP vs Life Expectancy (2007)")
```

---

## Summary

In this lesson, you learned how to:

- Create scatter plots  
- Compare distributions across categories  
- Create stacked and percent bar charts  
- Visualize time series  
- Build bubble charts  
- Explore multivariate relationships  

These techniques allow you to move from simple distributions to rich analytical storytelling using Plotly in R.
