---
title: "Bivariate and Multivariate Graphs with Plotly Express"
author: "SI Analytics"
format:
  html:
    css: global/style/si-analytics.css
    toc: true
    toc-location: left
    number-sections: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  error: false
---

## Introduction

This advanced module extends multivariate visualization into **analytical storytelling** using Plotly in R.

We move beyond exploration into:

- Regression overlays
- Trend lines
- Comparative subplots
- Multi-layered interpretation

All examples are designed for the RStudio environment.

---

## Libraries

```{r}
library(plotly)
library(dplyr)
library(gapminder)
```

---

# 1. Scatter Plot with Regression Line

We examine the relationship between vehicle weight and fuel efficiency.

```{r}
model <- lm(mpg ~ wt, data = mtcars)

mtcars$predicted <- predict(model)

p1 <- plot_ly(
  mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers",
  name = "Observed"
)

p2 <- plot_ly(
  mtcars,
  x = ~wt,
  y = ~predicted,
  type = "scatter",
  mode = "lines",
  name = "Regression Line"
)

subplot(p1, p2, nrows = 1) %>%
  layout(title = "Weight vs MPG with Regression Trend")
```

Interpretation:
As vehicle weight increases, fuel efficiency declines in a near-linear pattern.

---

# 2. Time Series with Trend Overlay

Nigeria population growth from Gapminder.

```{r}
nigeria <- gapminder %>%
  filter(country == "Nigeria")

trend_model <- lm(pop ~ year, data = nigeria)
nigeria$trend <- predict(trend_model)

plot_ly(nigeria, x = ~year, y = ~pop,
        type = "scatter", mode = "lines+markers",
        name = "Observed") %>%
  add_lines(y = ~trend, name = "Linear Trend") %>%
  layout(title = "Nigeria Population Growth with Trend Line")
```

This allows separation of signal (trend) from yearly variation.

---

# 3. Bubble Chart with Analytical Context

```{r}
g_2007 <- gapminder %>% filter(year == 2007)

plot_ly(
  g_2007,
  x = ~gdpPercap,
  y = ~lifeExp,
  size = ~pop,
  color = ~continent,
  type = "scatter",
  mode = "markers"
) %>%
  layout(title = "GDP vs Life Expectancy (2007)")
```

Storytelling angle:
- Wealthier nations generally live longer
- Population size introduces geopolitical scale
- Continental clustering reveals structural inequality

---

# 4. Faceted Subplots Using subplot()

We compare GDP vs Life Expectancy across continents.

```{r}
continents <- unique(g_2007$continent)

plots <- lapply(continents, function(cont) {
  subset_data <- g_2007 %>% filter(continent == cont)
  
  plot_ly(
    subset_data,
    x = ~gdpPercap,
    y = ~lifeExp,
    size = ~pop,
    type = "scatter",
    mode = "markers"
  ) %>%
    layout(title = cont)
})

subplot(plots, nrows = 2, shareX = TRUE, shareY = TRUE) %>%
  layout(title = "GDP vs Life Expectancy by Continent (2007)")
```

Faceting enables structured comparative analysis.

---

# 5. Comparative Distribution Subplots

Distribution of MPG by Cylinders and Gear.

```{r}
mtcars <- mtcars %>%
  mutate(cyl = factor(cyl), gear = factor(gear))

p_cyl <- plot_ly(mtcars, x = ~mpg, color = ~cyl,
                 type = "histogram") %>%
  layout(title = "MPG by Cylinders")

p_gear <- plot_ly(mtcars, x = ~mpg, color = ~gear,
                  type = "histogram") %>%
  layout(title = "MPG by Gear")

subplot(p_cyl, p_gear, nrows = 1, shareY = TRUE)
```

This reveals structural efficiency differences across engine configurations.

---

# Advanced Storytelling Principles

When adding regression lines or trend overlays:

1. Do not merely show correlation; interpret it.
2. Separate structural signal from noise.
3. Use subplots for comparison, not clutter.
4. Ensure axis consistency for valid comparison.
5. Let visuals support narrative, not replace it.

---

# Summary

In this advanced module, you learned how to:

- Add regression lines to scatter plots
- Overlay trend lines on time series
- Use bubble charts for multivariate insight
- Build faceted subplots using `subplot()`
- Structure visualizations for analytical storytelling

These techniques elevate visualization from descriptive to explanatory analytics.
