---
title: "Univariate Graphs with Plotly Express"
author: "SI Analytics"
format:
  html:
    css: ../global/style/si-analytics.css
    toc: true
    toc-location: left
    number-sections: false
execute:
  echo: true
  eval: false
  warning: false
  message: false
  error: false
---

## Intro

In this lesson, you will learn how to create univariate graphs using Plotly in **R** within the RStudio environment. 

Univariate graphs help us understand the distribution of a single variable, whether categorical or quantitative.

Let us get started.

## Learning Objectives

- Create histograms for quantitative variables using Plotly in R
- Create bar charts for categorical variables
- Create pie charts for categorical distributions
- Customize titles, axis labels, and colors

## Libraries

```{r}
library(plotly)
library(dplyr)
library(ggplot2)
```

---

# Quantitative Data

## Histogram

Histograms visualize the distribution of continuous variables.

We will use the built-in `mtcars` dataset.

```{r}
head(mtcars)
```

### Basic Histogram of MPG

```{r}
plot_ly(
  mtcars,
  x = ~mpg,
  type = "histogram"
)
```

::: {.callout-note title="Side Note"}
Plotly graphs are interactive. Hover over the bars to see frequencies. You can zoom and export the chart directly from the toolbar.
:::

---

## Customizing the Histogram

```{r}
plot_ly(
  mtcars,
  x = ~mpg,
  type = "histogram",
  marker = list(color = "#6a5acd")
) %>%
  layout(
    title = "Distribution of Miles Per Gallon",
    xaxis = list(title = "Miles Per Gallon (MPG)"),
    yaxis = list(title = "Frequency")
  )
```

---

::: {.callout-tip title="Practice"}

### Practice Q: Histogram of Horsepower

Create a histogram showing the distribution of horsepower (`hp`) in the `mtcars` dataset.  
Add:
- A title  
- Custom x-axis label  
- A hex color  

```{r}
# Your code here
```
:::

```{r}
#| include: false
# Solution
plot_ly(
  mtcars,
  x = ~hp,
  type = "histogram",
  marker = list(color = "#1f77b4")
) %>%
  layout(
    title = "Distribution of Horsepower",
    xaxis = list(title = "Horsepower"),
    yaxis = list(title = "Frequency")
  )
```
---

# Categorical Data

```{r}
mtcars <- mtcars %>%
  mutate(
    cyl = factor(cyl),
    gear = factor(gear)
  )
```

---

## Bar Chart

```{r}
plot_ly(
  mtcars,
  x = ~cyl,
  type = "histogram"
) %>%
  layout(
    title = "Distribution of Cylinders",
    xaxis = list(title = "Number of Cylinders"),
    yaxis = list(title = "Count")
  )
```

---

## Colored Bar Chart

```{r}
plot_ly(
  mtcars,
  x = ~cyl,
  type = "histogram",
  color = ~cyl,
  colors = c("#1f77b4", "#ff7f0e", "#2ca02c")
) %>%
  layout(
    title = "Distribution of Cylinders",
    showlegend = FALSE
  )
```

---

::: {.callout-tip title="Practice"}

### Practice Q: Distribution of Gears

Create a bar chart showing the distribution of `gear`.  
Add:
- Title  
- Custom colors  
- Remove legend  

```{r}
# Your code here
```
:::

```{r}
#| include: false
# Solution
plot_ly(
  mtcars,
  x = ~gear,
  type = "histogram",
  color = ~gear,
  colors = c("#4B0082", "#FF1493", "#20B2AA")
) %>%
  layout(
    title = "Distribution of Gears",
    showlegend = FALSE
  )
```
---

## Sorting Categories

```{r}
plot_ly(
  mtcars,
  x = ~cyl,
  type = "histogram"
) %>%
  layout(
    xaxis = list(categoryorder = "total descending")
  )
```

---

## Horizontal Bar Chart

```{r}
plot_ly(
  mtcars,
  y = ~cyl,
  type = "histogram"
) %>%
  layout(
    title = "Horizontal Distribution of Cylinders",
    xaxis = list(title = "Count"),
    yaxis = list(title = "Number of Cylinders")
  )
```

---

## Pie Chart

```{r}
plot_ly(
  mtcars,
  labels = ~cyl,
  type = "pie"
)
```

### Pie Chart with Labels

```{r}
plot_ly(
  mtcars,
  labels = ~cyl,
  type = "pie"
) %>%
  layout(showlegend = FALSE) %>%
  style(textposition = "inside", textinfo = "percent+label")
```

---

::: {.callout-tip title="Practice"}

### Practice Q: Pie Chart of Gears

Create a pie chart showing the distribution of `gear`.  
Include percentages and labels inside the slices.

```{r}
# Your code here
```
:::

```{r}
#| include: false
# Solution
plot_ly(
  mtcars,
  labels = ~gear,
  type = "pie"
) %>%
  layout(showlegend = FALSE) %>%
  style(textposition = "inside", textinfo = "percent+label")
```

---

# Summary

In this lesson, you learned how to:

- Create histograms for quantitative variables  
- Create bar charts for categorical variables  
- Sort categorical variables  
- Create horizontal bar charts  
- Create pie charts  
- Customize titles, labels, and colors  

You should now feel confident producing interactive univariate visualizations using Plotly in R within RStudio.
