---
title: "9  Univariate Graphs with Plotly in R"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    highlight: kate
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, echo = F, message = F, warning = F}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(rlang, tidyverse, knitr, here, reactable, gt, flextable, plotly, dplyr, )

## functions
source(here::here("global/functions/misc_functions.R"))

## default render
knitr::opts_chunk$set(class.source = "tgc-code-block", render = reactable_5_rows)

```

## 9.1 Intro

In this lesson, you will learn how to create **univariate graphs** using **Plotly in R**.\
Univariate graphs help us understand the distribution of a single variable, whether categorical or quantitative.

All examples are designed to run inside the **RStudio environment** using Quarto.

Let us get started.

------------------------------------------------------------------------

## 9.2 Learning Objectives

By the end of this lesson, you should be able to:

-   Create histograms for quantitative variables using Plotly in R
-   Create bar charts for categorical variables
-   Create pie charts for categorical distributions
-   Sort categorical variables in meaningful order
-   Customize titles, axis labels, and colors

------------------------------------------------------------------------

## 9.3 Setup

Load libraries:

```{r}
library(plotly)
library(dplyr)
```

------------------------------------------------------------------------

## 9.4 Quantitative Data

### 9.4.1 Histogram

Histograms visualize the distribution of continuous variables.

We will use the built-in `tips` dataset from plotly.

```{r}
data(tips, package = "plotly")
head(tips)
```

Create a basic histogram of tip amounts:

```{r}
fig <- plot_ly(
  tips,
  x = ~tip,
  type = "histogram"
)

fig
```

Hover over the bars to see frequencies. Plotly charts are interactive by default.

------------------------------------------------------------------------

### Customizing the Histogram

```{r}
fig <- plot_ly(
  tips,
  x = ~tip,
  type = "histogram",
  marker = list(color = "#6a5acd")
) %>%
  layout(
    title = "Distribution of Tips",
    xaxis = list(title = "Tip Amount ($)"),
    yaxis = list(title = "Frequency")
  )

fig
```

------------------------------------------------------------------------

### 9.4.2 Practice: Bird Strike Speed Histogram

```{r}
data(birdstrikes)

fig <- plot_ly(
  birdstrikes,
  x = ~Speed_IAS_in_knots,
  type = "histogram",
  marker = list(color = "#1f77b4")
) %>%
  layout(
    title = "Distribution of Aircraft Speed During Bird Strikes",
    xaxis = list(title = "Speed (Nautical Miles Per Hour)"),
    yaxis = list(title = "Frequency")
  )

fig
```

------------------------------------------------------------------------

## 9.5 Categorical Data

### 9.5.1 Bar Chart

Bar charts display the frequency of categorical variables.

```{r}
fig <- plot_ly(
  tips,
  x = ~sex,
  type = "histogram"
) %>%
  layout(
    title = "Distribution of Customers by Gender",
    xaxis = list(title = "Gender"),
    yaxis = list(title = "Count")
  )

fig
```

------------------------------------------------------------------------

### Adding Colors and Removing Legend

```{r}
fig <- plot_ly(
  tips,
  x = ~sex,
  type = "histogram",
  color = ~sex,
  colors = c("#1f77b4", "#ff7f0e")
) %>%
  layout(
    title = "Distribution of Customers by Gender",
    showlegend = FALSE
  )

fig
```

------------------------------------------------------------------------

### 9.5.2 Bird Strikes by Phase of Flight

```{r}
fig <- plot_ly(
  birdstrikes,
  x = ~When__Phase_of_flight,
  type = "histogram",
  color = ~When__Phase_of_flight
) %>%
  layout(
    title = "Bird Strikes by Phase of Flight",
    xaxis = list(title = "Phase of Flight"),
    yaxis = list(title = "Count"),
    showlegend = FALSE
  )

fig
```

------------------------------------------------------------------------

## 9.5.3 Sorting Categories by Frequency

```{r}
fig <- plot_ly(
  birdstrikes,
  x = ~Origin_State,
  type = "histogram"
) %>%
  layout(
    title = "Bird Strikes by Origin State (Descending)",
    xaxis = list(categoryorder = "total descending"),
    yaxis = list(title = "Count")
  )

fig
```

------------------------------------------------------------------------

## 9.5.4 Horizontal Bar Chart

When there are many categories, horizontal bar charts improve readability.

```{r}
fig <- plot_ly(
  birdstrikes,
  y = ~Origin_State,
  type = "histogram"
) %>%
  layout(
    title = "Horizontal Bar Chart of Origin State",
    yaxis = list(categoryorder = "total descending"),
    xaxis = list(title = "Count")
  )

fig
```

------------------------------------------------------------------------

## 9.5.5 Pie Chart

Pie charts display proportions of categorical variables.

```{r}
fig <- plot_ly(
  birdstrikes,
  labels = ~Wildlife__Size,
  type = "pie"
) %>%
  layout(
    title = "Distribution of Bird Strikes by Wildlife Size",
    showlegend = FALSE
  )

fig
```

------------------------------------------------------------------------

## 9.6 Summary

In this lesson, you learned how to create univariate graphs in R using Plotly:

-   Histograms for quantitative data
-   Bar charts for categorical data
-   Sorted and horizontal bar charts
-   Pie charts for proportions
-   Custom titles, labels, and colors

You should now feel confident producing interactive univariate visualizations in RStudio using Quarto.

See you in the next lesson.
